Novelty Detection Using Graphical Models for Semantic Room Classification

### 2011-05-17 semantic-nd/article.tex
Fixed Results section according to Andrzej comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
In order to verify the performance of the proposed thresholds a synthetic dataset
was generated. To keep it simple only information regarding direct properties of
a room were modelled and no structured knowledge such as room connectivity was taken
in account.
The synthetic distribution assumes that an independent and variable size set of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 7 different measured feature
types. Each feature type can be present more than once in (Eg.: room shape is
extracted from 2D laser scans in more than one position in the room).

The room categories were chosen to mimic as close as possible real features and
categories existent on reality. And they model different room categories with
different levels of detail. For example: 1 person office, 2 person office, hallway,
robot lab.

From the distribution, 100 labelled samples for 5 of the 11 room categories were
drawn to represent the known concepts and 1000 unlabelled samples were drawn from
all the room categories for learning the unconditional probability distribution.
Using those samples, factors were learnt for the graphs used to model the
conditional distribution and the independent unconditional distribution.
\autoref{fig:simple-experiment} shows the graph structure used for approximate the
trained conditional and unconditional distributions.
Its important to notice that graph $G$ used to model the known classes perfectly
mimics the real distribution, with the exception that the system is not aware of
all the room categories.

@@

% Describe how to obtain the 3 thresholds functions seen on the graphs.
Using the learned models $G$, $U$ and $I$ two thresholds were trained:
$P_G(x)/P_U(x)$ assuming a uniform unconditional distribution
and $P_G(x)/P_I(x)$ assuming an independent unconditional distribution.
Since the distribution is synthetic there is access to $P(x)$ and $P(x|concept)$
and so a perfect threshold function could also be created to test how far the
presented thresholds are from optimal.

\subsection{Probability ratio comparison}
%%% Results 1
% Show the threshold ratio is an optimal detector (if perfect information was available)
% Show that the thresholds are suitable functions for implementing a static threshold.
As first tryout, the performance of the novelty thresholds was plotted for a set
of 1000 samples taken out from the whole distribution (\autoref{fig:synthetic-roc}).
Those samples contain different number of sensed properties for each room, miming
the dynamic properties expected to see when implemented on a robot.

@@

The convex shape of the optimal threshold shows that the ratio between conditional
and unconditional probability is indeed an optimal detector and is suitable for
implementing a static threshold when the samples are taken from dynamic
distributions (eg.: some samples where there is only access to room size versus
samples where there is a lot of information on the room properties).

% Discuss importance on approximating unconditional probability.
Its also possible to see how important it is to estimating a correct unconditional
probability in order to obtain a correct novelty ordering on the inputs.
The assumption of an uniform unconditional probability has led to very poor results.
That is probably explained by semantic properties as the ones modelled to be highly
biased towards some values and understanding that bias plays an important step
to detect whether a given sensed value is a valuable clue on the room category
or not.

%%% Results 2
% Measure performance of the thresholds as more information becomes available.
\subsection{Performance changes with amount of available information}
In order to measure the performance impact as more semantic information becomes
available ROC curves were plotted for samples grouped by the number of sensed
semantic features.

@@

As possible to see as the system gains more semantic information it becomes easier
to detect novelty as the input space size increases and allows the several existing
classes to become more easily distinguished.

The performance of the independent threshold decreases as the number of sensed
features increases. This is easily explained by the fact that the graph $I$ is not
able to model the existent dependence between the features and that becomes obvious
as the number of features increases (Eg.: graph $I$ perfectly models $P(x)$ in the
case where only 1 feature is sensed).

The uniform threshold shows a very poor performance specially on small size samples
where its performs almost no better than random.
It performance increases as the size of sensed features increases but nonetheless
its always very small when compared to how optimal a threshold could be.
This shows that at least under semantic distributions as the one modelled with the
synthetic data assuming an uniform unconditional distribution leads to poor results.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
In order to verify the performance of the proposed threshold functions a synthetic dataset
was generated. To keep it simple only information regarding direct features of
a room were modelled and no structured knowledge such as room connectivity was taken
in account.
The synthetic distribution assumes that an independent and variable size set of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 7 different measured feature
types. Each feature type can be present more than once in (e.g.: room shape is
extracted from 2D laser scans in more than one position in the room).

The room categories were chosen to mimic as close as possible the real features and
categories existing in reality. And they model different room categories with
different levels of detail. For example: 1 person office, 2 person office, hallway,
robot lab.

From the distribution, 100 labelled samples for 5 of the 11 room categories were
drawn to represent the known concepts and 1000 unlabelled samples were drawn from
all the room categories for learning the unconditional probability distribution.
Using those samples, factors were learnt for the graphs used to model the
conditional distribution and the independent unconditional distribution.
\autoref{fig:simple-experiment} shows the graph structure used for approximate the
trained conditional and unconditional distributions.
Its important to notice that graph $G$ used to model the known classes when given
enough labelled data is able to exactly learn the conditional distribution as it
uses the same structure as the created synthetic distribution.

@@

% Describe how to obtain the 3 thresholds functions seen on the graphs.
Using the learned models $G$, $U$ and $I$, two thresholds were trained:
$P_G(x)/P_U(x)$ assuming a uniform unconditional distribution
and $P_G(x)/P_I(x)$ assuming an independent unconditional distribution.
Since the distribution is synthetic there is access to $P(x)$ and $P(x|concept)$
and a perfect threshold function could also be created to test how far the
presented thresholds are from optimal.

\subsection{Probability Ratio Comparison}
%%% Results 1
% Show the threshold ratio is an optimal detector (if perfect information was available)
% Show that the thresholds are suitable functions for implementing a static threshold.
As first tryout, the performance of the novelty threshold selection was plotted for a set
of 1000 samples taken out from the whole distribution (\autoref{fig:synthetic-roc}).
Those samples contain different number of sensed properties for each room, mimicking
the dynamic properties expected to see when implemented on a robot.

@@

The convex shape for the optimal threshold shows that the ratio between conditional
and unconditional probability is indeed an optimal detector and is suitable for
implementing a threshold when the samples are taken from dynamic
distributions (e.g.: some samples where there is only access to room size versus
samples where there is a lot of information about the room properties).

% Discuss importance on approximating unconditional probability.
Its also possible to see how important it is to estimate a correct unconditional
probability in order to obtain a correct novelty measure on the inputs.
The assumption of a uniform unconditional probability has led to very poor results.
That is probably explained by the semantic properties being highly
biased towards some values. And shows that bias plays an important step
in detecting whether a given sensed value is a valuable cue about the room category.

%%% Results 2
% Measure performance of the thresholds as more information becomes available.
\subsection{Performance Changes With Amount of Available Information}
In order to measure the performance impact as more semantic information becomes
available ROC curves were plotted for samples grouped by the number of sensed
semantic features.

@@

We can see that as the system gains more semantic information it becomes easier
to detect novelty. The input space size increases and allows the several existing
classes to become more easily distinguished.

The performance of the independent threshold decreases as the number of sensed
features increases. This is easily explained by the fact that the graph $I$ is not
able to model the existent dependence between the features. This becomes obvious
as the number of features increases (e.g.: graph $I$ perfectly models $P(x)$ in the
case where only 1 feature is sensed).

The uniform threshold shows a poor performance specially on small size samples
where its performs almost no better than random.
It performance increases as the size of sensed features increases but nonetheless
its very small when compared to how optimal a threshold could be.
